R Notebook

### 2021-12-14 src/main/r/numeric_vs_hpo_for_CRC.Rmd
include demographics in model

# preprocess: convert the HPO column (and the isnegated column) into three separate features:
# is_hpoterm_observed_for_H, is_hpoterm_observed_for_L and is_hpoterm_observed_for_N
# pass in the dataframe of lab tests of one LOINC, return a new dataframe with the three new features
create_hpo_features <- function(df_one_loinc){
  df_hpo_features <- df_one_loinc %>%
    mutate(dummy_value = 1) %>%
    pivot_wider(names_from = interpretation, names_prefix = 'is_hpoterm_observed_for_', values_from = dummy_value, values_fill = 0) %>%
    select(lab_result_numeric_value, starts_with('is_hpoterm_observed_for_'), outcome)

@@

lr_for_hpo_H <- function(df_train, df_test){
  model_continuous <- glm(formula = 'outcome ~ lab_result_numeric_value', data = df_train, family = binomial(link = 'logit'))
  model_hpo <- glm(formula = 'outcome ~ is_hpoterm_observed', data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), family = binomial(link = 'logit'))
  prediction_continuous <- predict(model_continuous, df_test, type = 'response')
  prediction_hpo <- predict(model_hpo, df_test %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), type = 'response')
  roc_continuous <- roc(df_test$outcome, prediction_continuous)
  roc_hpo <- roc(df_test$outcome, prediction_hpo)
  return (list(model_continuous, model_hpo, prediction_continuous, prediction_hpo, roc_continuous, roc_hpo))
}

lr_for_hpo_L <- function(df_train, df_test){
  model_continuous <- glm(formula = 'outcome ~ lab_result_numeric_value', data = df_train, family = binomial(link = 'logit'))
  model_hpo <- glm(formula = 'outcome ~ is_hpoterm_observed', data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), family = binomial(link = 'logit'))
  prediction_continuous <- predict(model_continuous, df_test, type = 'response')
  prediction_hpo <- predict(model_hpo, df_test %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), type = 'response')
  roc_continuous <- roc(df_test$outcome, prediction_continuous)
  roc_hpo <- roc(df_test$outcome, prediction_hpo)
  return (list(model_continuous, model_hpo, prediction_continuous, prediction_hpo, roc_continuous, roc_hpo))
}

@@

pipeline_compare_for_outcome_prediction <- function(df_one_loinc) {

@@

  roc_comparisons <- list()
  if ('H' %in% valid_lab_outcomes$interpretation){
    lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
    roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
    roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'H'))
    roc_comparisons$H <- roc_comparison
  }

  if ('L' %in% valid_lab_outcomes$interpretation){
    lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
    roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
    roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'L'))
    roc_comparisons$L <- roc_comparison
  }

@@

ggplot(matched_unique_ids_etc, aes(x = outcome, y = age)) +
  geom_violin()
```
We need to put back the identifiers of controls to the database in order to selectively retrieve their labs. 
```{r}
dbExecute(dsca, "create table if not exists matched_control_cohort (medical_record_number VARCHAR)")
dbExecute(dsca, "truncate table matched_control_cohort")

@@

# labs for controls
df_labs_and_outcomes_matched_controls <- df_matched_control_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
  as_tibble() %>%
  #filter(medical_record_number == '7007952')  %>%
  #mutate(lab_date = as.Date(lab_date)) %>%
  arrange(medical_record_number, loinc, desc(lab_date)) %>%
  group_by(medical_record_number, loinc) %>%
  mutate(sequence = row_number()) %>%
  ungroup() %>%
  filter(sequence == 1) %>%
  mutate(outcome = 'normal')

@@

>>>>

# preprocess: convert the HPO column (and the isnegated column) into three separate features:
# is_hpoterm_observed_for_H, is_hpoterm_observed_for_L and is_hpoterm_observed_for_N
# pass in the dataframe of lab tests of one LOINC, return a new dataframe with the three new features
create_hpo_features <- function(df_one_loinc){
  df_hpo_features <- df_one_loinc %>%
    mutate(dummy_value = 1) %>%
    pivot_wider(names_from = interpretation, names_prefix = 'is_hpoterm_observed_for_', values_from = dummy_value, values_fill = 0) %>%
    select(lab_result_numeric_value, starts_with('is_hpoterm_observed_for_'), any_of(c('age', 'gender', 'race')), outcome)

@@

lr_for_hpo_H <- function(df_train, df_test, include_demographics=FALSE){
  if (include_demographics){
    formula_continuous = 'outcome ~ lab_result_numeric_value + age + gender + race'
    formula_hpo = 'outcome ~ is_hpoterm_observed + age + gender + race'
  } else {
    formula_continuous = 'outcome ~ lab_result_numeric_value'
    formula_hpo = 'outcome ~ is_hpoterm_observed'
  }
  model_continuous <- glm(formula = formula_continuous, data = df_train, family = binomial(link = 'logit'))
  model_hpo <- glm(formula = formula_hpo, data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), family = binomial(link = 'logit'))
  prediction_continuous <- predict(model_continuous, df_test, type = 'response')
  prediction_hpo <- predict(model_hpo, df_test %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), type = 'response')
  roc_continuous <- roc(df_test$outcome, prediction_continuous)
  roc_hpo <- roc(df_test$outcome, prediction_hpo)
  return (list(model_continuous, model_hpo, prediction_continuous, prediction_hpo, roc_continuous, roc_hpo))
}

lr_for_hpo_L <- function(df_train, df_test, include_demographics=FALSE){
  if (include_demographics){
    formula_continuous = 'outcome ~ lab_result_numeric_value + age + gender + race'
    formula_hpo = 'outcome ~ is_hpoterm_observed + age + gender + race'
  } else {
    formula_continuous = 'outcome ~ lab_result_numeric_value'
    formula_hpo = 'outcome ~ is_hpoterm_observed'
  }

  model_continuous <- glm(formula = formula_continuous, data = df_train, family = binomial(link = 'logit'))
  model_hpo <- glm(formula = formula_hpo, data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), family = binomial(link = 'logit'))
  prediction_continuous <- predict(model_continuous, df_test, type = 'response')
  prediction_hpo <- predict(model_hpo, df_test %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), type = 'response')
  roc_continuous <- roc(df_test$outcome, prediction_continuous)
  roc_hpo <- roc(df_test$outcome, prediction_hpo)
  return (list(model_continuous, model_hpo, prediction_continuous, prediction_hpo, roc_continuous, roc_hpo))
}

@@

pipeline_compare_for_outcome_prediction <- function(df_one_loinc, include_demographics=FALSE) {

@@

  roc_comparisons <- list()
  if ('H' %in% valid_lab_outcomes$interpretation){
    lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test, include_demographics=include_demographics)
    roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
    roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'H'))
    roc_comparisons$H <- roc_comparison
  }

  if ('L' %in% valid_lab_outcomes$interpretation){
    lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test, include_demographics=include_demographics)
    roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
    roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'L'))
    roc_comparisons$L <- roc_comparison
  }

@@

plot_match_age <- ggplot(matched_unique_ids_etc, aes(x = outcome, y = age, fill = outcome)) +
  geom_violin(draw_quantiles = c(0.5)) +
  theme(legend.position = 'none')

plot_match_gender <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = gender)) +
  geom_bar(stat = 'count')

plot_match_race <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = race)) +
  geom_bar(stat = 'count')

plot_match <- gridExtra::arrangeGrob(plot_match_age, plot_match_gender, plot_match_race, layout_matrix = rbind(c(1,2), c(3,3)), ncol = 1)
plot(plot_match)
```
We need to put back the identifiers of controls to the database in order to selectively retrieve their labs. 
```{r}
dbExecute(dsca, "create table if not exists matched_control_cohort (medical_record_number VARCHAR)")
dbExecute(dsca, "truncate table matched_control_cohort")

@@

# labs for controls
df_labs_and_outcomes_matched_controls <- df_matched_control_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
  # my current control still contains CRC patients
  # TODO: further purify controls
  filter(!(medical_record_number %in% df_labs_and_outcomes_cases$medical_record_number)) %>%
  as_tibble() %>%
  #filter(medical_record_number == '7007952')  %>%
  #mutate(lab_date = as.Date(lab_date)) %>%
  arrange(medical_record_number, loinc, desc(lab_date)) %>%
  group_by(medical_record_number, loinc) %>%
  mutate(sequence = row_number()) %>%
  ungroup() %>%
  filter(sequence == 1) %>%
  mutate(outcome = 'normal')

@@

Combine demographics to see whether the AUC can be improved. 
```{r}
df_demographics_case_matched_control <- bind_rows(df_case_demographics %>% select(-min_diag_date), 
                                                  df_control_demographics %>% filter(medical_record_number %in% 
                                                                                       unique(df_matched_control_labs$medical_record_number)) %>%
                                                    filter(!(medical_record_number %in% df_case_demographics$medical_record_number)))

plan(multisession, workers=4)
start_time <- Sys.time()

crc_case_matched_controls_analysis_with_demographics <- df_labs_and_outcomes_case_matched_control %>%
  left_join(df_demographics_case_matched_control, by = 'medical_record_number') %>%
  mutate(age = as.duration(interval(date_of_birth, lab_date)) / dyears()) %>%
  filter(!(race %in% c('African American', 'American Indian or Alaska Native', 'Native Hawaiian or Other Pacific Islander'))) %>%
  # mutate(gender = as.factor(gender),
  #        race = as.factor(race)) %>% 
  # mutate(gender = relevel(gender, 'Male'),
  #        race = relevel(race, 'White')) %>%
  group_by(loinc) %>%
  nest() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, include_demographics=TRUE, .options = furrr_options(seed = T))) %>%
  select(-data) %>%
  unnest(result)

crc_case_matched_controls_analysis_with_demographics
```
